Indicators for Ecosystem Condition in Norway

### 2023-06-23 DRAFT_ndvi_wetland.Rmd
wetland ndvi continued

## filter out only wetland data
nin.wetland <- nin %>% 
  filter(hovedoekosystem %in% c('Vaatmark')) %>%
  mutate(id = identifikasjon_lokalid) %>%
  filter(validGeo) %>%
  drop_na(tilstand) %>%
  dplyr::select(id, hovedoekosystem, hovedtype, ninkartleggingsenheter, lokalitetskvalitet, tilstand, kartleggingsaar)

@@

nin.wetland <- nin.wetland %>% 
  mutate(area_meters_nin = st_area(nin.wetland)
  )

@@

# We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
model.wetland.train <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.wetland.train)
model.wetland.train <- StepBeta(model.wetland.train)
summary(model.wetland.train)
```

NDVI values do vary between:
- Regions
- Ecosystem types
- Condition classes

It is evident that all three variables matter, for ease of interpretation we run the model separately for every region
```{r}
model.wetland.train.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Northern Norway",])
model.wetland.train.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Central Norway",])
model.wetland.train.W <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Western Norway",])
model.wetland.train.E <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Eastern Norway",])
model.wetland.train.S <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Southern Norway",])

>>>>

## filter out only wetland data
nin.wetland <- nin %>% 
  filter(hovedoekosystem %in% c('Vaatmark')) %>%
  mutate(id = identifikasjon_lokalid) %>%
  filter(validGeo) %>%
  drop_na(tilstand) %>%
  dplyr::select(id, hovedoekosystem, hovedtype, ninkartleggingsenheter, lokalitetskvalitet, tilstand, ninbeskrivelsesvariabler, kartleggingsaar)

@@

nin.wetland <- nin.wetland %>% 
  mutate(area_meters_nin = st_area(nin.wetland)
  )
# check and edit the order of regions
levels(nin.wetland$region)
nin.wetland$region <- as.factor(nin.wetland$region)
levels(nin.wetland$region)
nin.wetland$region <- factor(nin.wetland$region, levels = c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway"))
levels(nin.wetland$region)

@@

# make an extra numeric condition variable
unique(SentinelNDVI.wetland.train$tilstand)
SentinelNDVI.wetland.train <- SentinelNDVI.wetland.train %>% mutate(tilstand_num = recode(tilstand, 
                                            "God" = '0',
                                            "Moderat" = '1',
                                            "Redusert" = '2',
                                            "Svært redusert" = '3'))
SentinelNDVI.wetland.train$tilstand_num <- as.numeric(SentinelNDVI.wetland.train$tilstand_num)

# We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
model.wetland.cond.Sent <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.wetland.train)
model.wetland.cond.Sent <- StepBeta(model.wetland.cond.Sent)
summary(model.wetland.cond.Sent)
```
NDVI values do vary between:
- Regions
- Ecosystem types
- Condition classes

It is evident that there are differences between regions. But since we are not interested in the statistics for region differences, we run separate models for every region for ease of interpretation.
```{r, include = F}
rm(model.wetland.cond.Sent)
```

```{r}
model.wetland.cond.Sent.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Northern Norway",])
model.wetland.cond.Sent.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Central Norway",])
model.wetland.cond.Sent.W <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Western Norway",])
model.wetland.cond.Sent.E <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Eastern Norway",])
model.wetland.cond.Sent.S <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Southern Norway",])
